Formalization of Abstract Rewriting

### 2025-02-23 Relations/Paper/Definitions.tex
call update

\begin{definition}
    \begin{description}
        \item[$R \models Inc$] \emph{$R$ is increasing} if there exists a ``size function'' $|-| : A \to \nat$ satisfying, for all $a, b \in A$,
        $Rab \to |a| < |b|$  
        \item[$a \in \CP$] \sacomment{Need a definition of cofinality here}
    \end{description}
\end{definition}

Having a "size function" that maps to the natural numbers is computationally taxing to satisfy. 
Futher, it takes us outside of pure relational theory. The following definitions are for properties which allow us to forgo $\Inc$, for further discussion see Subsection \ref{subsec:def}.
\begin{definition} \label{def:rp} Let $A$ and $R$ be as above, and let $a \in A$.
    \begin{description} 
        \item[$s : \nat \to A$] \emph{A sequence $s \in A$} is a function from the natural numbers to $A$.  
        \item[$s \in R-increasing$] \emph{A sequence $s \in A$ is $R$-increasing} if every term is $R$-related to its preceding term.
        \item[$a$ is an $s$-bound] if $s (i) \mstep a$ for all $i \in \nat$. 
        \item[$a \in \BP_R$] if all $R$-increasing sequence $s$ with $s (0) = a$ there exists an $s$-bound.  
        \item[$R \models \RPm$] \emph{$R$ has the weak recurrence property} if for every increasing sequence $s$ with bound $b$ there exists an $i \in \nat$ such that $b \mstep s (i)$.
        \item[$R \models \RP$] \emph{$R$ has the recurrence property} if every bounded, increasing sequence $s$ with an $s$-bound, there exists an $i \in \nat$ such that 
        $s (i) \in \MF$. See Definition \ref{def:mf}. 
    \end{description}
\end{definition}
In \terese $\mathsf{Ind}$ is used to denote $\BP$.

@@

\subsection{Finitely branching (\texttt{FinitelyBranching.agda})}
\sacomment{This property is what makes a lot of the classical properties valid which otherwise wouldn't be. Valid in most interesting rewriting systems. Most 
programming languages will satisfy finite branching.}
\begin{definition} Let $A$ and $R$ be as above, and let $a \in A$.
    \begin{description}
        \item[$a \in \FB$] \emph{$a$ is finitely branching} if there is a finite amount of  $b \in A$ such that $a \rstep b$. 
    \end{description}
\end{definition}

>>>>

\begin{definition} \hfill
    \begin{description}
        \item[$R \models Inc$] \emph{$R$ is increasing} if there exists a ``size function'' $|-| : A \to \nat$ satisfying, for all $a, b \in A$,
        $Rab \to |a| < |b|$  
    \end{description}
\end{definition}

Having a size function that maps to the natural numbers is computationally taxing to satisfy. 
Further, it takes us outside of purely relational theory. The following definitions are for properties which allow us to forgo $\Inc$, for further discussion see Subsection \ref{subsec:def}.
\begin{definition} \label{def:rp} Let $A$ and $R$ be as above, and let $a \in A$.
    \begin{description} 
        \item[$s : \nat \to A$] \emph{A sequence $s \in A$} is a function from the natural numbers to $A$.  
        \item[$s \in R-increasing$] \emph{A sequence $s \in A$ is $R$-increasing} if every term is $R$-related to its preceding term.
        \item[$a$ is an $s$-bound] if $s (i) \mstep a$ for all $i \in \nat$. 
        \item[$a \in \BP_R$] if all $R$-increasing sequence $s$ with $s (0) = a$ there exists an $s$-bound.  
        \item[$R \models \RPm$] \emph{$R$ has the weak recurrence property} if for every increasing sequence $s$ with bound $b$ there exists an $i \in \nat$ such that $b \mstep s (i)$.

        \item[$R \models \RP$] \emph{$R$ has the recurrence property} if every bounded, increasing sequence $s$ with an $s$-bound, there exists an $i \in \nat$ such that 
        $s (i) \mstep x \implies x \mstep s (i)$ for all $x\in A$. See Definition \ref{def:mf}. 
    \end{description}
\end{definition}
In \terese $\mathsf{Ind}$ is used to denote $\BP$.

@@

\subsection{Finitely branching (\texttt{FinitelyBranching.agda})}

\begin{definition} Let $A$ and $R$ be as above, and let $a \in A$.
    \begin{description}
        \item[$a \in \FB$] \emph{$a$ is finitely branching} if there is a finitely many $b \in A$ such that $a \rstep b$.  
    \end{description}
\end{definition}
Our Agda formulation of the above definition makes use of the List type constructor. 
\verb|FB a = Σ[ xs ∈ List A ] (∀ b → R b a → b ∈List xs)|

$\FB$ is a property that plays an important role in our discussion of well-foundedness in Section \ref{sec:Well-foundedness}.

### 2025-02-23 Relations/Paper/Formalization.tex
call update

The following is a summary of the differences in definition we have used as mentioned in Section \ref{sec:Definitions}. 
\begin{itemize}
    \item $\SN$ : We make use of the well-founded defintion, rather than the finite reduction sequence definition.
    \item $\NPe$ : We use the equivalent $\NP$ definition. 
    \item $\Inc$ : We use the more general property $\RP$.      
\end{itemize}

As already mentioned, the $\Inc$ property was one we wanted to move away from. We saw that $\Inc$ was being used only in Theorem 1.2.3 
and that the crucial facet of $\Inc$ for the proof of the theorem was that every sequence has a bound \sacomment{is this the correct characterization?}. 
So we established that the properties in Definition \ref{def:rp} were sufficient for the proving the theorems. This in turn led us to examining 
a generalization of a normal form, the minimal form. We explore this further in Section \ref{sec:Implications}.

@@

\sacomment{This is a point where it may be useful to examine MF earlier. However, the alternative is to discuss the generalization of NL in the section on implications.}

We show in Section \ref{sec:Implications} a generalization of Newman's Lemma. \sacomment{Delete this comment when that generalization is written.}

@@

($ii$) is formalized in:
\begin{itemize}
    \item \verb|ii : R isWN × R isUN → R isCR|
\end{itemize}
In our formalization we show that $\gWN \land \gUNto \implies \gCR$ is true and so provide a generalization of ($ii$) as $\gUN \implies \gUNto$.

@@

\subsubsection{Theorem 1.2.3}
There are four parts to this theorem. 
\begin{description}
    \item[($i$)] $\gWN \and \gUN \implies \gBP$
    \item[($ii$)] $\gBP \land \gInc \implies \gSN$  
    \item[($iii$)] $\gWCR \land \gWN \land \gInc \implies \gSN$
    \item[($iv$)] $\gCR \iff \gCR$ ($\implies$ only holds if the ARS is countable.)
\end{description}
($i$) is formalized in: 
\begin{itemize}
    \item \verb|i : R isWN → R isUN → R isBP|
    \item \verb|i→ : R isWN → R isUN→ → R isBP|
\end{itemize}
In $\mathtt{i\to}$ we again provide a generalization by using $\gUNto$ rather than $\gUN$.

@@

The reverse implication of ($iv$) is formalized in: 
\begin{itemize}
    \item \verb|iv : R isCP → R isCR|
\end{itemize}
We did not formalize $\gCR \to gCP$ as it requires the ARS to be countable \sacomment{More to say?}.

>>>>

The following is a summary of the differences in definition we have used as mentioned in Section \ref{sec:Definitions}. 
\begin{itemize}
    \item $\SN$ : We make use of the well-founded definition, rather than the finite reduction sequence definition.
    \item $\NPe$ : We use the equivalent $\NP$ definition. 
    \item $\Inc$ : We use the more general property $\RP$.      
\end{itemize}

As already mentioned, the $\Inc$ property was one we wanted to move away from. We saw that $\Inc$ was being used only in Theorem 1.2.3 
and that the crucial facet of $\Inc$ for the proof of the theorem was that no $R$-increasing 
sequence can have an upper bound.
We established that the properties $\RP$ and $\RPm$ were sufficient for the proving the theorems. This in turn led us to 
examine the condition in $\RP$ more closely. This condition provides a generalization of the notion of normal form, 
which we analyze in detail in Section \ref{sec:Implications}. \sacomment{Try and get this para to flow better.}

@@

We show in Subsection \ref{subsec:newnewman} a generalization of Newman's Lemma using $\MF$.

@@

($ii$) is formalized in:
\begin{itemize}
    \item \verb|ii : R isWN × R isUN → R isCR|
\end{itemize}
\sacomment{Make below comment tidier}
Note that we have provided a proof with the weaker hypothesis of $\gUNto$.

@@

\subsubsection{Theorem 1.2.3}
The following definition is required for this theorem.
\begin{definition} Let $a \in A$.
    \begin{description}
        \item[$a \in CP_R$] \emph{$a$ has the cofinality property} if there exists a 
        (not necessarily strictly) increasing sequence starting from $a$ such that every
        reduct of $a$ reduces to some element of the sequence.
    \end{description}
\end{definition}

There are four parts to this theorem. 
\begin{description}
    \item[($i$)] $\gWN \and \gUN \implies \gBP$
    \item[($ii$)] $\gBP \land \gInc \implies \gSN$  
    \item[($iii$)] $\gWCR \land \gWN \land \gInc \implies \gSN$
    \item[($iv$)] $\gCR \iff \gCP$ ($\gCR \implies \gCP$ only holds if the ARS is countable.)
\end{description}
($i$) is formalized in: 
\begin{itemize}
    \item \verb|i : R isWN → R isUN → R isBP|
    \item \verb|i→ : R isWN → R isUN→ → R isBP|
\end{itemize}
In $\mathtt{i\to}$ we again provide a generalization by using $\gUNto$ rather than $\gUN$.

@@

\sacomment{neeed to define CP}
The reverse implication of ($iv$) is formalized in: 
\begin{itemize}
    \item \verb|iv : R isCP → R isCR|
\end{itemize}
We did not formalize $\gCR \to gCP$ as it requires the ARS to be countable \sacomment{More to say?}.

### 2025-02-23 Relations/Paper/Implications.tex
call update

\sacomment{Need to relate SM and SMseq. Probably define SMseq and metion 
we can only prove this for SMseq.}
\begin{proposition}\label{prop:SMRP}
    $SM \leftrightarrow RP \land BP$
\end{proposition}
\begin{proof}
    See,\\ \verb|RP∧BP→SMseq : R isRP → R isBP → ∀ {x : A} → SMseq R x| \\ \verb|RisSMseq→RisRP : (∀ {x : A} → SMseq R x) → R isRP|
    \\ \verb|RisSMseq→RisBP : (∀ {x : A} → SMseq R x) → R isBP| in \texttt{ARS-Implications.agda}.
\end{proof}

@@

The implications in this diagram are formalized in \texttt{ARS-Implications.agda}. The classical property of being minimally decidable is 
required to complete the implcation from $\SN$ to $\WN$, as can be seen in our formalization

>>>>

\sacomment{Need to relate SM and SMseq. Probably define SMseq and mention 
we can only prove this for SMseq.}
\begin{proposition}\label{prop:SMRP}
    $SM \leftrightarrow RP \land BP$
\end{proposition}
\begin{proof}
    See,\\ \verb|RP∧BP→SMseq : R isRP → R isBP → ∀ {x : A} → SMseq R x| \\ \verb|RisSMseq→RisRP : (∀ {x : A} → SMseq R x) → R isRP|
    \\ \verb|RisSMseq→RisBP : (∀ {x : A} → SMseq R x) → R isBP| in \texttt{ARS-Implications.agda}.
\end{proof}

@@

The implications in this diagram are formalized in \texttt{ARS-Implications.agda}. The classical property of being minimally decidable is 
required to complete the implication from $\SN$ to $\WN$, as can be seen in our formalization
